---
title: "Exercise set 2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Learning goals

After completing these tasks, you'll be able to read in csv-formatted data, use different functions to look at the data, create simple plots of the data, to save the plots you created. New concepts: data structures, factors, NA, writing to file, plot elements

## 2.1. More data structures
Q2.1.1 Create three vectors called x,y,z, each of them numeric and 4 elements long.

### Matrices
Q2.1.2 Join these three vectors together using the command `cbind()`.

Q2.1.3 Print out (onto the screen) the third value on the second row using square bracket notation.

Q2.1.4 Use the function `matrix()` to create a 3-by-3 matrix filled with uniformly distributed random numbers (runif). 

Tip: you need to create 9 random numbers and then give the resulting vector as a parameter to matrix function, defining the number of rows and/or columns.

Q2.1.5 Print out the 2-by-2 matrix which consists of rows 2 and 3 and columns 2 and 3 of the larger matrix in Q2.1.4.

### Lists
Q2.1.6 Create a list whose elements are your matrices from Q2.1.2 and Q2.1.4.

Q2.1.7 Through the list, access the first element of the first matrix.

## 2.2 Working with data
### Loading data
Q2.2.1 Download the exercise data to your local computer.

Q2.2.2 Load the csv file with the command `data <- read.csv([yourfilelocation], sep=';')`.

### First looks at your data
Q2.2.3 Take an initial look at the data by printing out the first five rows with `head`.

Q2.2.4 Print out the last 10 rows with `tail`. You need to explicitly give an argument to get 10 instead of 5, can you find out from the help what that parameter is called?

Q2.2.5 Get some summary statistics with `summary`.

Q2.2.6 Load the package 'psych' and get a different kind of summary statistics representation with `describe`. Note: You should already have psych installed if you did the last set of exercises.

### Indexing a data frame
Q2.2.7 Use square brackets to access the 7th row.

Q2.2.8 Use $ indexing to access the co_Gender column.

Q2.2.9 Can you spot what is wrong with the data in Q2.2.7 and Q2.2.8?

Q2.2.10 Convert categorical variables to factors with with factor(). For gender, 1=female, 2=male. For education level, 1=low, 2=middle, 3=high.

Q2.2.11 Run summary again, see what has changed in summary for the columns you made into factors.

Q2.2.12 Use square bracket notation and a vector of column names to access age, gender, education level and Healthy_pattern for all rows.

### Plotting

Q2.2.13 Run `plot(data[,c('Age_calculated','Healthy_pattern')])`, look at the resulting plot, and then run `plot(data[,c('co_EDUC_fam_clas','Healthy_pattern')])`. You used the same command, why do the plots look different? 

Tip: You can use the arrows at the top of plot viewer to view earlier plots you've made. We will go over combining multiple plots in the same figure next week.

Q2.2.14 Let's build a histogram of the use of fresh vegetables in this sample! Start by running `hist(data$Freshveg)`

Q2.2.15 Looks like we need more bars to get a better look, add parameter `breaks=20`. You can play around with the number to find one which feels good to you.

Q2.2.16 Try adding x and y axis labels with `xlab="Fresh vegetable consumption frequency (times per week)"` and `ylab="Number of participants"`

Q2.2.17 Add title to your plot with the parameter `main`

Q2.2.18 Add a colour to the bars with `col`. 
Tip: You can get a list of acceptable colour name strings with `colours()` (Note: Checking the colours is something it would be fine to just type into the console and not include in the script - you're just looking up what options you have and once you've decided, you do not need to run the check again). You can also define your own rgb colour with rgb()

Q2.2.19 Once you're happy with your plot, save it as a pdf-file. 

Tip: Remember to put your pdf() command before you build the plot and run dev.off() after you're done until the console says null device.

Q2.2.20 Run the following code
```{r eval=FALSE}
with(data, hist(Freshveg[co_Gender=='male'], 
                col=rgb(0,0,1,0.5), 
                breaks=0:35, 
                freq=F, 
                xlim=c(0,35),
                xlab="Fresh vegetable consumption frequency (times per week)", 
                ylab="Number of participants", 
                main="Fresh vegetable use"))
with(data, hist(Freshveg[co_Gender=='female'], 
                col=rgb(1,0,0,0.5), 
                breaks=0:35, 
                freq=F, 
                add=T))
legend("topright", legend=c("Girls", "Boys"),
       col=c(rgb(1,0,0,0.5), rgb(0,0,1,0.5)), lwd=10)
```
Do you understand what this code does? What does the parameter freq control? What about the parameter add?

Tip: You can try to comment out lines or change the input values to see what changes

Q2.2.21 Make a scatter plot which shows BMI as a function of age. You can also practice adding axis labels, a title, and changing other graphical parameters if you'd like. Save the resulting plot.

Tip: If you want to color by categorical variable, in scatter plots it is easier than with histograms. You just need to add `col=as.integer(data$co_EDUC_fam_clas)` (or any other factor vector) as an argument. 

## 2.3 If-else structure

Q2.3.1 Run the following loop
```{r eval=FALSE}
# an example loop for going over 20 first rows of the data set
for(row in 1:20){ # if you wanted to go over the whole data set, you would set the end to be nrow(data)
  if (is.na(data$Chocolate[row])){
    cat(paste0("We do not know if subject ", data$ID_number[row], " eats chocolate.\n"))
  } else if (data$Chocolate[row]==0){
    cat(paste0("Subject ", data$ID_number[row], " does not eat chocolate.\n"))
  } else {
    cat(paste0("Subject ", data$ID_number[row], " eats chocolate.\n"))
  }
}
```

Q2.3.2 Edit the loop above so that if the subject eats chocolate and their weekly chocolate consumption is more than the average amount (function mean, remember to use na.rm=T) it will output "Subject X eats more than average amount of chocolate per week" and if they eat less than the average amount, it will ouput "Subject X eats less than average amount of chocolate per week"

Q2.3.3 Create a new loop which goes over each row of your data frame. For each row, the loop will print "This is subject number " + the subject's ID number

Q2.3.4 Continue building the loop by adding another sentence to the output. If the subject is male, the loop should print "He is X years old." and if the subject is female, it should print "She is X years old." If the subject's gender is not known (is.na), it should print out "They are X years old."

Tip: Control for NA first, otherwise you will get an error when the loop encounters its first NA.

