---
title: "Webinar 2 tasks"
output: html_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Make loops work

This exercise is a somewhat silly example of constructing loops, but it can hopefully illustrate a more general use case where you need to load in multiple files (that have been named in a systematic way), manipulate the data in some way and then save the results into one data structure.

Your task is to take the code below, and edit the middle part to be a loop instead of copy&paste of the same code. Make the loop run over the 19 different files (named ouput_1.jpg, ouput_2.jpg etc) from the zip file. Then you can run the last lines of code as-is to convert the images into gif and to show that gif.

[Get a .zip of the files associated with the task here](./images/loop_images.zip)

```{r, eval=FALSE}


library(magick) #NB: remember to install magick if you don't already have it!

### edit from here to the next bit with ###

# change this file location to point to the folder where you extracted the images from the zip-file
filelocation <- '/Users/juusu53/Documents/projects/opetus/R_for_scientists/course_materials/' 

# reading in image from a file
new_image_1 <- image_read(paste0(filelocation, 'ouput_1.jpg'))
# adding annotations to the image
new_image_1 <- image_annotate(new_image_1, 'You did it!!', color = "white", size = 50, gravity='north')
# checking to see how it looks like
print(new_image_1)

new_image_2 <- image_read(paste0(filelocation, 'ouput_2.jpg'))
new_image_2 <- image_annotate(new_image_2, 'You did it!!', color = "white", size = 50, gravity='north')
print(new_image_2)

new_image_3 <- image_read(paste0(filelocation, 'ouput_3.jpg'))
new_image_3 <- image_annotate(new_image_3, 'You did it!!', color = "white", size = 50, gravity='north')
print(new_image_3)

# combine all of the images into one vector
img_list <- c(new_image_1, new_image_2, new_image_3)

# Tip: when changing the code to a loop, you should first initialise an empty vector BEFORE the loop like this: img_list <- vector()
# then, instead of doing one big combine at the end, on every iteration of the loop, you should do this: img_list <- c(img_list, new_image)
# this will append (add to the end) the newly loaded image so that the list gets longer on every iteration of the loop

### see if it worked! ###
# (you should not need to edit anything below this point)

# join the images together
img_joined <- image_join(img_list)

# animate at 10 frames per second
img_animated <- image_animate(img_joined, fps = 10)

# view animated image
img_animated

```